A Transactional Memory System for Parallel Python

### 2014-06-08 talk/dls2014/paper/paper.tex
Results of a final pass through the paper

The standard model employed in these systems is fairly
straightforward: there can be multiple threads, but each acquires the
GIL, then performs a single instruction, and subsequently releases the
lock.  While this setup limits the benefits due to parallelism,
the use of a GIL also provides
some crucial guarantees. Since this lock is always acquired while
executing bytecode instructions and may be released only in-between
such instructions, the GIL ensures perfect isolation between multiple
threads for a series of instructions, which are executed atomically.
In addition, the GIL provides the application with a sequential
consistency model~\cite{lamport79}.

A global interpreter lock assures the atomic execution of individual
instructions but offers the programmer no help in managing concurrent
data structures. Many parallel programs must maintain invariants
involving multiple objects (or even multiple fields of a single
object).  So to enable Python programs for parallel execution, we need
a way to identify which blocks of data must be executed atomically and
hardware or software support to effect this atomic execution.

\emph{Atomic blocks}~\cite{tim03,tim05} allow dynamic languages to
indicate the units of computation that must be executed atomically to
ensure correct concurrent execution.  Atomic blocks map nicely to
transactional memory, which in turn provides a convenient mechanism to replace
the GIL.  Using transactions to enclose multiple bytecode
instructions, we can get the very same semantics as provided by the
GIL while possibly executing several transactions in parallel. And by
exposing these interpreter-level transactions to the application, we
avoid the well-known problems caused by locks as applications use
atomic blocks instead of explicit acquire/release operations on
locks~\cite{christopher10,victor11,shan08}.

@@

This paper makes the following contributions:
\vspace{-\topsep}               % XXX:HACK:TODO
\begin{itemize}
\item We introduce a new STM system that performs well for dynamic
  language interpreters.
\item We integrate the system closely with a garbage collector
  (GC) to lower the overhead of STM.
\item This new STM system is used to replace the GIL in one
  implementation of Python and is then evaluated.
\end{itemize}

@@

We implement and evaluate our system for the Python language. Python
is a dynamic programming language that was designed with GIL semantics
in mind. Its reference implementation, CPython~\cite{cpython}, uses a
GIL. Over the years, Python added multiple ways to provide concurrency and
parallelism to its applications. We want to highlight two of them,
namely \emph{threading} and \emph{multiprocessing}.

\emph{Threading} employs operating system (OS) threads to provide
concurrency. It is, however, limited by the GIL and thus does not
provide parallelism\footnote{At this point we should mention that it is indeed
possible to run external functions written in C instead of Python in
parallel. Our work focuses on Python itself and ignores this aspect as
it requires writing in a different language.}. Because the GIL is only
seen as an implementation detail of the interpreter, applications are
required to coordinate concurrent accesses to data structures using
conventional methods -- locks being the most common way.

@@

We focus on the \emph{threading} approach. This requires us to remove
the GIL from the interpreter to run code in parallel on multiple
threads. One approach to this is fine-grained locking instead of a
single global lock. Jython~\cite{webjython} and
IronPython~\cite{ironpython} follow this approach. Fine-grained
locking is, however, not a \emph{direct} replacement for the GIL. It
requires multiple locks in multiple places, not only in a central
location. We on the other side follow the direct approach of using TM
instead of the GIL. There, GIL acquisition and release operations
conceptually directly map to starting and committing transactions.

@@

In summary, translating an $\%gs:SO$ to a physical address is a
two-step process: First the memory segmentation feature of the CPU
constructs a linear address. Then, this LA gets mapped by the MMU to
some place in the physical memory. This makes the $SO$ a valid reference
to an object in all segments, automatically resolving either to a
shared or a private version.

@@

Our TM system is designed as a library that covers all aspects around
transactions and object memory management. It is designed for
object-oriented dynamic language VMs as a replacement for the GIL.
The library exposes the functions in Listing~\ref{lst:api} as its
interface for use by a dynamic language interpreter.

@@

The job of the write barrier is twofold: it serves as a write barrier
for the garbage collector (GC) and the STM system. These two jobs are
depicted in Figure~\ref{fig:obj_states}. The GC needs to keep track of
objects that may contain references to young, movable objects. These
objects need to be traced during minor collections to update
references to moved objects. Young objects automatically get traced if
the GC ascertains that they survive the collection by being reachable
from traced objects. The STM system needs to record objects in the
write set, acquire write locks, and privatise their pages
(copy-on-write). It only needs to do this for old objects from
previous transactions, as the others are not known to other
transactions anyway.

The \textbf{fast path} of the write barrier is very simple. We only
need to check for the flag \lstinline!WRITE_BARRIER! in the object's
header through its $SO$ and call the slow path if it is set. This flag
is set either if the object is old and comes from an earlier
transaction, or if there was a minor collection which will add the
flag again on all objects, including new overflow objects. The flag
is never set on young, freshly allocated objects. This fast path
covers all cases for the GC and for the STM system (see
Listing~\ref{lst:w_fast}).

@@

The \textbf{slow path} is shown in Listing~\ref{lst:w_slow}.
First comes the \emph{GC part}: In any case, the object will be added
to the list of objects that need tracing in the next minor collection
(\lstinline!to_trace!). Any reference we write to this object will
be kept alive and up-to-date during minor collections.

@@

For the \emph{STM part}, we first perform a read barrier on the
object. We then try to acquire its write lock. \lstinline!write_locks!
is a simple, \emph{global} array of bytes that is indexed with the
$SO$ of the object divided by 16. Note, ``global'' here really means
it is a single array with data for all segments, there is no address
translation going on to access its elements contrary to e.g.\ the
read markers array.  If we already own the lock, we are done.
If someone else owns the lock, we perform  write-write contention
management that aborts either us or the current owner of the
object.  If we succeed in acquiring the lock using an atomic
\lstinline!cmp_and_swap!, we need to add the object to the write set
(a simple list called \lstinline!modified_old_objects!)  and privatise
all pages belonging to it (copy-on-write).

@@

Committing a transaction needs a bit more work. First, we synchronise
all threads so that the committing one is the only one running and all
the others are waiting in safe points. We then go through the write
set (\lstinline!modified_old_objects!)  and check the corresponding
\lstinline!read_markers!  in other threads~/ segments. If we detect a
read-write conflict, we perform contention management to either abort us or
the other transaction, or to simply wait a bit (see Section~\ref{subsub:contentionmanagement}).

After verifying that there are no conflicts anymore, we copy all our
changes done to the objects in the write set to all other segments,
including the sharing-segment. This is safe since we synchronised all
threads. We also need to push overflow objects generated by minor
collections to other segments, since they may reside partially in
private pages. At that point we also get a new
\lstinline!overflow_number! by increasing a global one, so that it
stays globally unique for each transaction. Increasing the
\lstinline!read_version!  is then done at the start of a new
transaction.

\subsection{Thread Synchronisation}

@@

We evaluate our system in a Python interpreter called
PyPy~\cite{pypy} (version 2.3). PyPy is an implementation of an
interpreter for the Python language. It has a special focus on speed
and provides a just-in-time (JIT) compiler to speed up applications
running on top of it. We compare between normal PyPy
using a GIL and a PyPy with STM, as well as to CPython.
\begin{description}
\item[CPython] (version 2.7.6) is the reference implementation of the Python
  language. It is the most widely used interpreter for this language.
  The implementation uses a GIL for synchronisation in multi-threaded
  execution and it does not feature a JIT compiler.
\end{description}

@@

For each point in the plots, we took 5 measurements and report the
arithmetic mean with the standard deviation as error bars. Average
speedup numbers are calculated using the geometric mean. For the JIT
benchmarks, we first let the JIT warm up by doing a few additional
iterations to get a single measurement.

@@

\subsection{Non-JIT Benchmarks}
First we run our benchmarks on three different interpreters: CPython
(GIL), and PyPy with STM and with the GIL (both without the JIT). The
results are shown in Figure~\ref{fig:performance-nojit}.

@@

We see that generally, the group of embarrassingly parallel benchmarks
scales best. (There is a notable performance stability problem in the
\emph{richards} benchmark. This is a bug we will fix in the future.)
The other three benchmarks scale barely or not at all with the number of
threads. The reason for this is likely again the conflicts in the
latter group. Since the JIT accelerates all code but not the STM
overhead, we do more work per transaction. This increases the
likelihood of conflicts between them and therefore limits scalability
even more than in the no-JIT benchmarks.

@@

% STM needs good performance:
The solution presented here is based on a software TM system.  As the
STM is a vital part of the interpreter, good performance is
essential. We present an interesting way to combine the existing
memory segmentation and virtual memory features of current CPUs to
lower the overhead of STM. This software implementation reduces
dramatically the cost of transactional memory and at the same time
provides an opportunity to integrate the STM system with garbage
collection.  This implementation leverages the processor's memory
segmentation hardware, which has lately played only a minor role in
our operating system.  Unfortunately, processor architects seem to
undervalue this hardware resource -- in the evolution from the x86 to
the x86{-}64 architecture, 4 out of 6 segment registers have been
removed. This development potentially threatens the STM strategy
outlined here and we want to remind the architecture community that
this unit can support interesting implementations -- a viewpoint that
has also been expressed by other researchers (see also
\cite{bennet10}).

>>>>

The standard model employed in these systems is fairly
straightforward: there can be multiple threads, but each acquires the
GIL, then performs one (or several) instruction, and subsequently releases the
lock.  While this setup limits the benefits due to parallelism,
the use of a GIL also provides
some crucial guarantees. Since this lock is always acquired while
executing bytecode instructions and may be released only in-between
such instructions, the GIL ensures perfect isolation between multiple
threads for a series of instructions, which are executed atomically.
In addition, the GIL provides the application with a sequential
consistency model~\cite{lamport79}.

A global interpreter lock assures the atomic execution of individual
instructions but offers the programmer no help in managing concurrent
data structures. Many parallel programs must maintain invariants
involving multiple objects (or even multiple fields of a single
object).  So to enable Python programs for parallel execution, we need
a way to identify which blocks of data must be executed atomically and
hardware or software support to effect this atomic execution.
This is traditionally done with locks.

\emph{Atomic blocks}~\cite{tim03,tim05} would allow dynamic languages to
indicate the units of computation that must be executed atomically to
ensure correct concurrent execution.  Atomic blocks map nicely to
transactional memory, which in turn provides a convenient mechanism to replace
the GIL.  Using transactions to enclose multiple bytecode
instructions, we can get the very same semantics as provided by the
GIL while possibly executing several transactions in parallel. And by
exposing these interpreter-level transactions to the application, we
avoid the well-known problems caused by locks as applications use
atomic blocks instead of explicit acquire/release operations on
locks~\cite{christopher10,victor11,shan08}.

@@

This paper makes the following contributions:
%\vspace{-\topsep}               % XXX:HACK:TODO
\begin{itemize}
\item We introduce a new STM system that performs well for dynamic
  language interpreters.
\item We integrate the system closely with a garbage collector
  (GC) to lower the overhead of STM.
\item This new STM system is used to replace the GIL in one
  implementation of Python and is then evaluated.
\end{itemize}

@@

We implement and evaluate our system for the Python language. Python
is a dynamic programming language that was originally designed without
concurrency in mind.
Its reference implementation, CPython~\cite{cpython}, uses a
GIL. Over the years, Python added multiple ways to provide concurrency and
parallelism to its applications. We want to highlight two of them,
namely \emph{threading} and \emph{multiprocessing}.

\emph{Threading} employs operating system (OS) threads to provide
concurrency. It is, however, limited by the GIL and thus does not
provide parallelism\footnote{At this point we should mention that it is indeed
possible to run external functions written in C instead of Python in
parallel. Our work focuses on Python itself and ignores this aspect as
it requires writing in a different language.}.  Moreover, the GIL can
only ensure correctness of the interpreter itself: applications are
required to coordinate concurrent accesses to data structures using
conventional methods -- locks being the most common way.

@@

We focus on the \emph{threading} approach. This requires us to remove
the GIL from the interpreter to run code in parallel on multiple
threads. One approach to this is fine-grained locking instead of a
single global lock. Jython~\cite{webjython} and
IronPython~\cite{ironpython} follow this approach. Fine-grained
locking is, however, not a \emph{direct} replacement for the GIL. It
requires multiple locks in multiple places, not only in a central
location. We on the other hand follow the direct approach of using TM
instead of the GIL. There, GIL acquisition and release operations
conceptually directly map to starting and committing transactions.

@@

In summary, translating a $\%gs:SO$ to a physical address is a
two-step process: First the memory segmentation feature of the CPU
constructs a linear address. Then, this LA gets mapped by the MMU to
some place in the physical memory. This makes the $SO$ a valid reference
to an object in all segments, automatically resolving either to a
shared or a private version.

@@

Our TM system is designed as a C library that covers all aspects around
transactions and object memory management. It is designed for
object-oriented dynamic language VMs as a replacement for the GIL.
The library exposes the functions in Listing~\ref{lst:api} as its
interface for use by a dynamic language interpreter.

@@

The job of the write barrier is twofold: it serves as a write barrier
for the garbage collector (GC) and the STM system. These two jobs are
depicted in Figure~\ref{fig:obj_states}. The GC needs to keep track of
objects that may contain references to young, movable objects. These
objects need to be traced during minor collections to update
references to moved objects. Young objects automatically get traced if
the GC ascertains that they survive the collection by being reachable
from traced objects. So far, this is how generational GCs generally
work. But the STM system also needs to record objects in the
write set, acquire write locks, and privatise their pages
(copy-on-write). It only needs to do this for old objects from
previous transactions, as any other object is not known to other
transactions anyway.

The \textbf{fast path} of the write barrier is very simple. We only
need to check for the flag \lstinline!WRITE_BARRIER! in the object's
header through its $SO$ and call the slow path if it is set. This flag
is set either if the object is old and comes from an earlier
transaction, or if there was a minor collection (in this case, the flag
is added again on all objects, including new overflow objects). The flag
is never set on young, freshly allocated objects. This fast path
covers all cases for the GC and for the STM system (see
Listing~\ref{lst:w_fast}).

@@

The \textbf{slow path} is shown in Listing~\ref{lst:w_slow}.  For
each object, this code is called at most once per minor collection.
First comes the \emph{GC part}: In any case, the object will be added
to the list of objects that need tracing in the next minor collection
(\lstinline!to_trace!). Any reference we write to this object will
be kept alive and up-to-date during minor collections.

@@

For the \emph{STM part}, we first perform a read barrier on the
object. We then try to acquire its write lock. \lstinline!write_locks!
is a simple, \emph{global} array of bytes that is indexed with the
$SO$ of the object divided by 16. Note, ``global'' here really means
it is a single array with data for all segments, there is no address
translation going on to access its elements contrary to e.g.\ the
read markers array.  If we already own the lock, we are done.
If someone else owns the lock, we perform  write-write contention
management that aborts either us or the current owner of the
object.  If the lock is not owned so far, and if we succeed in
acquiring it using an atomic
\lstinline!cmp_and_swap!, we need to add the object to the write set
(a simple list called \lstinline!modified_old_objects!).  We then
privatise the page or pages where it resides (copy-on-write).

@@

Committing a transaction needs a bit more work. First, we synchronise
all threads so that the committing one is the only one running and all
the others are waiting in safe points.\footnote{This design choice is
important to support our very cheap barriers. It is a trade-off that
works best if the number of concurrent threads is small.
See section~\ref{subsub:sync} for how we mitigate the costs.}
We then go through the write
set (\lstinline!modified_old_objects!)  and check the corresponding
\lstinline!read_markers!  in other threads~/ segments. If we detect a
read-write conflict, we perform contention management to either abort us or
the other transaction, or to simply wait a bit (see Section~\ref{subsub:contentionmanagement}).

After verifying that there are no conflicts anymore, we copy all our
changes (i.e.\ objects in our write set) to all other segments,
including the sharing-segment. This is safe since we synchronised all
threads. We also need to push overflow objects generated by minor
collections to other segments, since they may reside partially in
private pages. At that point we also get a new
\lstinline!overflow_number! by increasing a global counter, so that it
stays globally unique for each transaction. Increasing the (local)
\lstinline!read_version!  is then done at the start of a new
transaction.

\subsection{Thread Synchronisation\label{subsub:sync}}

@@

We evaluate our system in a Python interpreter called
PyPy~\cite{pypy}.
\begin{description}
\item[PyPy] (version 2.3) is an implementation of an
  interpreter for the Python language. It has a special focus on speed
  and provides a just-in-time (JIT) compiler to speed up applications
  running on top of it. We compare between normal PyPy
  using a GIL and a PyPy with STM, as well as to CPython.
\item[CPython] (version 2.7.6) is the reference implementation of the Python
  language. It is the most widely used interpreter for this language.
  The implementation uses a GIL for synchronisation in multi-threaded
  execution and it does not feature a JIT compiler.
\end{description}

@@

For each point in the plots, we took 5 measurements and report the
arithmetic mean with the standard deviation as error bars. Average
speedup numbers are calculated using the geometric mean. For the JIT
benchmarks, we first let the JIT warm up by doing a few additional
iterations before the actual measurement.

@@

\subsection{Non-JIT Benchmarks}
First we run our benchmarks on three different interpreters: CPython
(GIL), PyPy with STM, and PyPy with the GIL (all without the JIT). The
results are shown in Figure~\ref{fig:performance-nojit}.

@@

We see that generally, the group of embarrassingly parallel benchmarks
scales best. (There is a notable performance stability problem in the
\emph{richards} benchmark. This is a bug we will fix in the future.)
The other three benchmarks scale barely or not at all with the number of
threads. The reason for this is likely again the conflicts in the
latter group. Right now, because the code runs much more quickly with
the JIT than without, it has the effect of making the transactions
logically longer.  This increases the
likelihood of conflicts between them and therefore limits scalability
even more than in the no-JIT benchmarks.

@@

% STM needs good performance:
The solution presented here is based on a software TM system.  As the
STM is a vital part of the interpreter, good performance is
essential. We present an interesting way to combine the existing
memory segmentation and virtual memory features of current CPUs to
lower the overhead of STM. This software implementation reduces
dramatically the cost of transactional memory and at the same time
provides an opportunity to integrate the STM system with garbage
collection.  This implementation leverages the processor's memory
segmentation hardware, which has lately played only a minor role in
our operating systems.  Unfortunately, processor architects seem to
undervalue this hardware resource -- in the evolution from the x86 to
the x86{-}64 architecture, 4 out of 6 segment registers have been
removed. This development potentially threatens the STM strategy
outlined here and we want to remind the architecture community that
this unit can support interesting implementations -- a viewpoint that
has also been expressed by other researchers (see also
\cite{bennet10}).
